---
title: "WDDA FS 2026: Guide for Problem Set 7"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: false
    number_sections: true
    latex_engine: xelatex
lang: en
---

# Introduction

This guide provides detailed explanations for the exercises in WDDA FS 2026 Problem Set 7. The set focuses on **inference in regression models** -- that is, confidence intervals, significance tests, and prediction intervals. We learn how reliable estimated coefficients and predictions are and how to test hypotheses about model parameters.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(readxl)
library(ggplot2)
library(dplyr)
library(mosaic)
```

# Exercise 1: Advertising -- Confidence intervals and bootstrap

**Task:**
Analyse the advertising dataset and compare classical and bootstrap confidence intervals for the coefficients of simple and multiple regressions. Examine the significance of the predictors and perform a model comparison.

## Step 1: Read data and fit a simple model (a)

```{r}
Advertising <- read_excel("../../data/WDDA_07.xlsx", sheet = "Advertising")
head(Advertising)

mod_tv <- lm(sales ~ TV, data = Advertising)
summary(mod_tv)
```

**Interpretation:**

- **Intercept ($\beta_0 \approx 7.03$):** expected sales when `TV = 0`.
- **TV coefficient ($\beta_1 \approx 0.048$):** for each additional 1\$ of TV spending, sales increase by about 0.048 units on average.
- Both coefficients are highly significant ($p < 0.001$).

## Step 2: Classical confidence interval (b)

```{r}
confint(mod_tv)
```

**Interpretation:** The 95\% confidence interval for the TV coefficient is roughly between 0.042 and 0.053 -- i.e. we are 95\% confident that the true effect of TV spending on sales lies in this range.

## Step 3: Bootstrap confidence interval (c)

```{r}
set.seed(123)
simcoef <- do(5000) * coef(lm(sales ~ TV, data = resample(Advertising)))

CI_intercept <- quantile(simcoef$Intercept, c(0.025, 0.975))
CI_tv <- quantile(simcoef$TV, c(0.025, 0.975))

cat("Bootstrap 95% CI for Intercept:", round(CI_intercept, 3), "\n")
cat("Bootstrap 95% CI for TV:       ", round(CI_tv, 3), "\n")
```

**Comparison:** The bootstrap intervals agree closely with the classical intervals. This is expected when the classical assumptions (normality of residuals, constant variance) are reasonably well met.

## Step 4: Visualising the bootstrap distribution (d)

```{r, fig.width=7, fig.height=4}
par(mfrow = c(1, 2))
hist(simcoef$TV, main = "Bootstrap distribution: TV", xlab = "TV coefficient",
     col = "lightgray", border = "white", breaks = 25)
abline(v = quantile(simcoef$TV, c(0.025, 0.975)), col = "red", lty = 2, lwd = 2)

hist(simcoef$Intercept, main = "Bootstrap distribution: Intercept", xlab = "Intercept",
     col = "lightgray", border = "white", breaks = 25)
abline(v = quantile(simcoef$Intercept, c(0.025, 0.975)), col = "red", lty = 2, lwd = 2)
par(mfrow = c(1, 1))
```

**Assessment:** Both distributions are approximately bell-shaped (central limit theorem). The red lines mark the 95\% confidence bounds.

## Step 5: Multiple regression (e)

```{r}
mod_full <- lm(sales ~ TV + radio + newspaper, data = Advertising)
summary(mod_full)
```

**Interpretation:**

- `TV` and `radio` are highly significant ($p < 0.001$).
- `newspaper` is **not significant** ($p \approx 0.86$) -- its effect is statistically indistinguishable from zero.
- The F-test ($p < 2.2 \cdot 10^{-16}$) shows that the model overall explains sales significantly.

## Step 6: Model comparison via ANOVA F-test (f)

```{r}
mod_red <- lm(sales ~ TV + radio, data = Advertising)
anova(mod_red, mod_full)
```

**Interpretation:**

- The F-test compares the reduced model (`TV + radio`) with the full model.
- $p > 0.05$ means: `newspaper` does not contribute significant additional explanatory power.
- **Conclusion:** we can drop `newspaper` from the model without losing meaningful explanatory power.

## Step 7: Prediction with confidence and prediction intervals (g)

```{r}
new_values <- data.frame(TV = 200, radio = 30)

ci_mean <- predict(mod_red, newdata = new_values, interval = "confidence")
pi_obs  <- predict(mod_red, newdata = new_values, interval = "prediction")

cat("Confidence interval for mean sales:\n")
print(ci_mean)
cat("\nPrediction interval for an individual observation:\n")
print(pi_obs)
```

**Interpretation:**

- The **confidence interval** (`lwr`, `upr` in `ci_mean`) refers to the **expected value** of sales at the given predictor values -- i.e. the mean over many hypothetical observations.
- The **prediction interval** is wider because it additionally accounts for the spread of **individual** observations around the regression surface.

# Exercise 2: Diamond Rings -- Inference and prediction

**Task:**
For the Diamond Rings dataset, investigate the relationship between weight and price. Compute confidence and prediction intervals and test the significance of the slope.

## Step 1: Data and model (a)

```{r}
diamonds <- read_excel("../../data/WDDA_07.xlsx", sheet = "Diamonds Rings")
names(diamonds) <- c("weight", "price")
head(diamonds)

mod_dr <- lm(price ~ weight, data = diamonds)
summary(mod_dr)
```

**Interpretation:**

- **Slope ($\approx 3721$):** for each additional carat, the price increases on average by about 3721 SGD.
- The t-value for `weight` is large, $p$ is extremely small -- the effect is highly significant.

## Step 2: Confidence interval for the slope (b)

```{r}
confint(mod_dr)
```

**Interpretation:** The 95\% confidence interval for the slope is narrow and clearly above zero -- the relationship is statistically well established.

## Step 3: Bootstrap confidence interval (c)

```{r}
set.seed(42)
simcoef_dr <- do(5000) * coef(lm(price ~ weight, data = resample(diamonds)))

CI_weight <- quantile(simcoef_dr$weight, c(0.025, 0.975))
cat("Bootstrap 95% CI for slope (weight):", round(CI_weight, 1), "\n")
```

**Comparison:** Bootstrap and classical intervals are very similar in this case.

## Step 4: Prediction for a new ring (d)

```{r}
newr <- data.frame(weight = 0.25)

ci_mean <- predict(mod_dr, newdata = newr, interval = "confidence")
pi_obs  <- predict(mod_dr, newdata = newr, interval = "prediction")

cat("Mean expected price (95% CI):\n")
print(ci_mean)
cat("\nPrediction interval for a single ring (95% PI):\n")
print(pi_obs)
```

**Interpretation:** The prediction interval is wider than the confidence interval because it also accounts for the individual variability around the regression line.

## Step 5: Visualisation with intervals (e)

```{r, fig.width=8, fig.height=5}
grid <- data.frame(weight = seq(min(diamonds$weight), max(diamonds$weight), length.out = 100))
ci <- predict(mod_dr, newdata = grid, interval = "confidence")
pi <- predict(mod_dr, newdata = grid, interval = "prediction")

plot(diamonds$weight, diamonds$price, pch = 16, col = "gray50",
     xlab = "Weight (carats)", ylab = "Price (SGD)",
     main = "Regression with 95% CI and PI")
lines(grid$weight, ci[, "fit"], col = "blue", lwd = 2)
lines(grid$weight, ci[, "lwr"], col = "blue", lty = 2)
lines(grid$weight, ci[, "upr"], col = "blue", lty = 2)
lines(grid$weight, pi[, "lwr"], col = "red",  lty = 3)
lines(grid$weight, pi[, "upr"], col = "red",  lty = 3)
legend("topleft", legend = c("Fit", "95% CI", "95% PI"),
       col = c("blue", "blue", "red"), lty = c(1, 2, 3), lwd = c(2, 1, 1))
```

**Assessment:** The PI band (red) is clearly wider than the CI band (blue). Both are narrowest at the centre of the data and widen towards the edges.

# Exercise 3: SAT -- Multiple regression and the F-test

**Task:**
For the SAT dataset, examine which educational variables best explain the average SAT score of a US state. Perform model comparisons using an F-test.

## Step 1: Read and inspect the data (a)

```{r}
sat <- read_excel("../../data/WDDA_07.xlsx", sheet = "SAT")
head(sat)
summary(sat[, c("expend", "ratio", "salary", "frac", "sat")])
```

**Description:**

- `expend`: per-pupil education spending (in 1000\$)
- `ratio`: pupil-teacher ratio
- `salary`: average teacher salary (in 1000\$)
- `frac`: fraction of students taking the SAT (in \%)
- `sat`: average SAT score

## Step 2: Simple model (b)

```{r}
mod1 <- lm(sat ~ expend, data = sat)
summary(mod1)
```

**Surprise:** The slope on `expend` is **negative**! Higher education spending is associated with lower SAT scores? This looks paradoxical -- a hint at a confounding variable.

## Step 3: Multiple regression (c)

```{r}
mod2 <- lm(sat ~ expend + ratio + salary + frac, data = sat)
summary(mod2)
```

**Interpretation:**

- After controlling for `frac` (fraction of students participating), **the sign of many coefficients flips**.
- `frac` is highly significant and negative: states where more students take the SAT have lower average scores (self-selection).
- `expend` and `salary` now have positive signs -- the true effect of additional resources is actually positive once we control for the selection effect.

## Step 4: F-test of the overall model (d)

```{r}
f <- summary(mod2)$fstatistic
cat("F statistic:", round(f["value"], 2),
    "  (df1 =", f["numdf"], ", df2 =", f["dendf"], ")\n")
```

**Interpretation:** The F-test checks the global null hypothesis that **all** slopes are zero. A high F-value with a small $p$ rejects this hypothesis -- at least one predictor explains significant variance.

## Step 5: ANOVA comparison of nested models (e)

```{r}
mod_small <- lm(sat ~ frac, data = sat)
mod_large <- lm(sat ~ expend + ratio + salary + frac, data = sat)
anova(mod_small, mod_large)
```

**Interpretation:**

- Compares the reduced model (only `frac`) with the full model.
- A significant $p$-value would mean that the extra predictors jointly contribute significantly to the explanation.
- If $p > 0.05$: the extra variables provide no significant gain -- `frac` alone captures most of the variation.

## Step 6: Prediction for a hypothetical state (f)

```{r}
hypothetical <- data.frame(expend = 6, ratio = 16, salary = 35, frac = 30)
pred <- predict(mod_large, newdata = hypothetical, interval = "prediction")
print(pred)
```

**Interpretation:** Expected SAT score plus 95\% prediction interval for a state with the specified characteristics.

# Summary

This problem set has shown:

1. **Classical and bootstrap confidence intervals** yield very similar results when model assumptions hold. Bootstrap is more robust when assumptions are violated.
2. **t-tests** check individual coefficients, **F-tests** check the overall model or model comparisons.
3. **Confidence vs. prediction intervals**: the CI refers to the expected value, the PI to a single new observation -- the PI is always wider.
4. **Multiple regression** allows estimating effects of individual predictors controlling for other variables -- important for avoiding confounding (SAT example).
5. **ANOVA F-tests** compare nested models and help with model selection: simpler models are preferred unless they are significantly worse.
