---
title: "R Code Examples from WDDA Lecture 6 (Chapter 6)"
subtitle: "WDDA -- Data Management and Data Analysis"
author: "Prof. Dr. Ulrich Matter"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
    latex_engine: xelatex
    md_extensions: "+lists_without_preceding_blankline"
    includes:
      in_header: ../style/notes_preamble.tex
lang: en
---

# Introduction

This document collects all R code examples from the slides of WDDA FS2026 Lecture 6 (Chapter 6). The focus is on multiple linear regression, an extension of simple linear regression to several predictors.

Multiple linear regression is a statistical method that models the relationship between a dependent variable and several independent variables. In contrast to simple linear regression, which uses only a single predictor, multiple regression allows us to investigate more complex relationships.

The central concepts in this area are:

1. **Multiple regression with several predictors**: How do we model the influence of several variables?
2. **Residual analysis and goodness of fit**: How well does the model fit the data?
3. **Prediction and error margin**: How accurate are the model's predictions?
4. **Non-linear and polynomial regression**: How do we model non-linear relationships?
5. **Model comparison**: How do we compare different regression models?

# Preparation

```{r vorbereitung-sichtbar, eval=FALSE}
# Load packages
library(scatterplot3d)
library(rgl)
library(car)
library(readxl)

# Load data
data <- read_excel("data/WDDA_06.xlsx", sheet = "Advertising")

# Make variables directly addressable
attach(data)
```

```{r vorbereitung, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
# Load packages
library(scatterplot3d)
library(rgl)
library(car)
library(readxl)

# Load data
data <- read_excel("../../data/WDDA_06.xlsx", sheet = "Advertising")

# Make variables directly addressable
attach(data)

# Inspect the data structure
str(data)
```

**R explanation**:

- The function `library()` loads an installed package into the current R session.
- The package `scatterplot3d` enables the creation of 3D scatterplots.
- The package `rgl` provides interactive 3D visualisations.
- The package `car` (Companion to Applied Regression) contains functions for regression analysis.
- With `read_excel()` from the package `readxl` you can import Excel files.
- The function `attach()` makes the variables of a data frame directly addressable, without having to prefix the data frame name.
- The function `str()` shows the structure of an object, including data types and dimensions.

**Note on data paths**: In R Markdown files, relative paths are interpreted from the location of the file, hence the path `"../data/WDDA_06.xlsx"`. In ordinary R scripts the path would be specified relative to the working directory.

# Investigating the Explanatory Variables

Before we begin with the multiple regression, we examine the relationships between the variables.

```{r untersuchung_variablen, eval=TRUE}
# Correlations
cor_tv <- cor(sales, TV)
cor_radio <- cor(sales, radio)
cat("Correlation (TV vs. Sales):", round(cor_tv, 4), "\n")
cat("Correlation (Radio vs. Sales):", round(cor_radio, 4), "\n\n")

# Correlation matrix
cor_matrix <- cor(data[, c("TV", "radio", "newspaper", "sales")])
print(round(cor_matrix, 4))

# 3D visualisation (TV, Radio, Sales)
scatterplot3d(x = TV, y = radio, z = sales, scale.y = 0.9, angle = 30,
              main = "3D Scatterplot: TV, Radio and Sales",
              xlab = "TV (in 1000 USD)", ylab = "Radio (in 1000 USD)", zlab = "Sales (in 1000 USD)")

# Pairwise scatterplots
pairs(data[, c("TV", "radio", "newspaper", "sales")],
      main = "Pairwise Scatterplots")
```

```{r dynamischer_3d_plot, eval=FALSE}
# Dynamic 3D scatterplot
# Note: this function opens an interactive window
scatter3d(x = TV, z = radio, y = sales, surface = FALSE)
```

**Statistical explanation**:

- The **correlation** is a measure of the linear relationship between two variables and lies between -1 and +1.
- A **correlation matrix** shows the pairwise correlations between several variables.
- **3D scatterplots** allow the visualisation of the relationship between three variables.
- **Pairwise scatterplots** show the relationships between all pairs of variables in a dataset.
- In multiple regression it is important to understand the relationships between the predictors, since multicollinearity (strong correlations between predictors) can complicate the interpretation of the coefficients.

**R explanation**:

- The function `cor()` computes the Pearson correlation between two vectors or a correlation matrix for a data frame.
- The function `scatterplot3d()` from the package of the same name creates a 3D scatterplot.
- The function `scatter3d()` from the package `car` creates an interactive 3D scatterplot.
- The function `pairs()` creates a matrix of pairwise scatterplots.

# Multiple Linear Regression (TV and Radio)

We now run a multiple regression with two predictors: TV and Radio.

```{r multiple_regression, eval=TRUE}
# Model with two predictors
md2 <- lm(sales ~ TV + radio)
summary(md2)

# Regression equation
B <- coef(md2)
cat("Regression equation:\n")
cat("sales ≈", round(B[1], 3), "+", round(B[2], 3), "* TV +", round(B[3], 3), "* radio\n\n")

# Comparison with simple regression (TV only)
md1 <- lm(sales ~ TV)
summary(md1)

# Comparison of the coefficients
cat("Coefficient for TV (simple regression):", round(coef(md1)[2], 4), "\n")
cat("Coefficient for TV (multiple regression):", round(coef(md2)[2], 4), "\n\n")

# Interpretation of the coefficients
cat("Interpretation of the coefficients (multiple regression):\n")
cat("- Intercept:", round(B[1], 3),
    "→ expected sales when TV = 0 and radio = 0 (in thousands of USD)\n")
cat("- TV coefficient:", round(B[2], 4),
    "→ expected increase in sales for +1 unit of TV, holding radio constant\n")
cat("- Radio coefficient:", round(B[3], 4),
    "→ expected increase in sales for +1 unit of radio, holding TV constant\n")
```

**Statistical explanation**:

- **Multiple linear regression** models the relationship between a dependent variable and several independent variables: y = β₀ + β₁x₁ + β₂x₂ + ... + βₚxₚ + ε
- The **regression coefficients** (β₁, β₂, ...) give the influence of each independent variable on the dependent variable when all other variables are held constant.
- The **intercept** (β₀) is the expected value of the dependent variable when all independent variables are zero.
- The **interpretation of the coefficients** in multiple regression differs from that in simple regression:
  - In simple regression, the coefficient gives the total effect of the variable.
  - In multiple regression, the coefficient gives the partial effect of the variable when all other variables are held constant.
- The coefficients can differ between simple and multiple regression when the predictors are correlated.

**R explanation**:

- The function `lm()` (linear model) fits a linear regression model to the data.
- The formula `sales ~ TV + radio` in `lm()` specifies that `sales` is the dependent variable and `TV` and `radio` are the independent variables.
- The function `summary()` provides a detailed summary of the regression model, including coefficients, standard errors, t-values, p-values and R².
- The function `coef()` extracts the coefficients from a regression model.

# Prediction with the Model

With the regression model we can make predictions for new values of the independent variables.

```{r vorhersage, eval=TRUE}
# Prediction with error margin
RSE <- summary(md2)$sigma
tv_new <- 230.1
radio_new <- 37.8
predicted <- predict(md2, newdata = data.frame(TV = tv_new, radio = radio_new))
pred_interval <- predicted + c(-1, 1) * 2 * RSE
cat("Forecast for TV =", tv_new, "and Radio =", radio_new, ":", round(predicted, 2), "\n")
cat("95% prediction interval:", round(pred_interval[1], 2), "to", round(pred_interval[2], 2), "\n\n")

# Exact prediction interval
pred_interval_exact <- predict(md2, newdata = data.frame(TV = tv_new, radio = radio_new),
                              interval = "prediction", level = 0.95)
cat("Exact 95% prediction interval:",
    round(pred_interval_exact[2], 2), "to",
    round(pred_interval_exact[3], 2), "\n")

# Confidence interval for the expected value
conf_interval <- predict(md2, newdata = data.frame(TV = tv_new, radio = radio_new),
                        interval = "confidence", level = 0.95)
cat("95% confidence interval for the expected value:",
    round(conf_interval[2], 2), "to",
    round(conf_interval[3], 2), "\n")
```

**Statistical explanation**:

- The **prediction** (forecast) is the predicted value of the dependent variable for given values of the independent variables: ŷ = β₀ + β₁x₁ + β₂x₂ + ... + βₚxₚ
- The **residual standard error (RSE)** is a measure of the typical size of the residuals and has the same unit as the dependent variable.
- The **prediction interval** gives the range in which a single new value will lie with a given probability.
- The **confidence interval for the expected value** gives the range in which the true expected value lies with a given probability.
- Prediction intervals are wider than confidence intervals for the expected value, because they additionally take into account the variation of individual observations around the expected value.

**R explanation**:

- The function `predict()` computes predicted values for a regression model.
- The parameter `newdata` in `predict()` specifies a data frame with the values of the predictors for which predictions are to be made.
- The parameter `interval` in `predict()` specifies whether confidence or prediction intervals are to be computed.
- The parameter `level` in `predict()` specifies the confidence level.
- The expression `summary(md2)$sigma` extracts the RSE from the regression object.

# Regression Surface (Plane of Best Fit)

In multiple regression with two predictors we can visualise the regression surface in 3D space.

```{r regression-surface, eval=FALSE}
# Visualisation of the regression surface
# Note: this function opens an interactive window
scatter3d(x = TV, z = radio, y = sales, fit = "linear", residuals = TRUE, grid = TRUE)
```

```{r regression-surface_statisch, eval=TRUE}
# Static visualisation of the regression surface
s3d <- scatterplot3d(TV, radio, sales,
                    main = "Regression Surface (Plane of Best Fit)",
                    xlab = "TV", ylab = "Radio", zlab = "Sales",
                    pch = 16, highlight.3d = TRUE, angle = 45)

# Add the regression surface
my.lm <- lm(sales ~ TV + radio)
s3d$plane3d(my.lm, lty.box = "solid")
```

**Statistical explanation**:

- In **multiple regression with two predictors**, the relationship is represented by a plane in three-dimensional space: z = β₀ + β₁x + β₂y
- This plane is called the **regression surface** (plane of best fit).
- The regression surface is chosen so that the sum of the squared vertical distances of the data points to the surface is minimised.
- The **residuals** are the vertical distances of the data points to the regression surface.
- Visualising the regression surface helps to understand the joint influence of the two predictors on the dependent variable.

**R explanation**:

- The function `scatter3d()` from the package `car` creates an interactive 3D scatterplot with a regression surface.
- The parameter `fit = "linear"` in `scatter3d()` specifies that a linear regression surface should be fitted.
- The parameter `residuals = TRUE` in `scatter3d()` displays the residuals as vertical lines.
- The function `scatterplot3d()` creates a static 3D scatterplot.
- The method `plane3d()` adds a regression surface to a `scatterplot3d` object.

# Goodness of Fit and Explanatory Power

We now examine how well the model explains the variation in the data.

```{r model-fit, eval=TRUE}
# Calculation of TSS, RSS and R²
TSS <- sum((sales - mean(sales))^2)
RSS <- sum((sales - predict(md2))^2)
r_squared <- 1 - (RSS / TSS)

cat("Coefficient of determination (R²):", round(r_squared, 4), "\n")
cat("TV and Radio explain", round(r_squared * 100, 2), "% of the variation in Sales.\n\n")

# Comparison with R² from summary()
r_squared_summary <- summary(md2)$r.squared
cat("R² from summary():", round(r_squared_summary, 4), "\n")

# Comparison with simple regression (TV only)
r_squared_tv <- summary(md1)$r.squared
cat("R² (TV only):", round(r_squared_tv, 4), "\n")
cat("R² (TV + Radio):", round(r_squared_summary, 4), "\n")
cat("Increase in R² from adding Radio:",
    round((r_squared_summary - r_squared_tv) * 100, 2), "percentage points\n")
```

**Statistical explanation**:

- The **coefficient of determination R²** gives the proportion of the variation in the dependent variable that is explained by the regression model.
- R² lies between 0 and 1:
  - R² = 0: the model explains none of the variation in the data.
  - R² = 1: the model explains all of the variation in the data.
- R² can be computed in several ways:
  - R² = 1 - (RSS / TSS)
  - R² = ESS / TSS
- A higher R² means a better fit of the model to the data.
- Adding further predictors can only increase R² or leave it unchanged, never decrease it.
- This can lead to overfitting if too many predictors are added.

**R explanation**:

- The function `predict()` without further parameters computes the predicted values for the observations in the training dataset.
- The expression `summary(md2)$r.squared` extracts the R² value from the regression object.
- R² is computed directly via the formula R² = 1 - (RSS / TSS).

# Residual Analysis

The analysis of the residuals is an important step for checking the assumptions of linear regression.

```{r residuenanalyse, eval=TRUE}
# Residual plots
hist(resid(md2), main = "Histogram of the Residuals", xlab = "Residuals")
plot(resid(md2) ~ TV, main = "Residuals vs. TV", xlab = "TV", ylab = "Residuals")
abline(h = 0, col = "gray")
plot(resid(md2) ~ radio, main = "Residuals vs. Radio", xlab = "Radio", ylab = "Residuals")
abline(h = 0, col = "gray")

# Residuals vs. fitted values
plot(resid(md2) ~ predict(md2), main = "Residuals vs. Fitted Values", xlab = "Prediction", ylab = "Residuals")
abline(h = 0, col = "gray")

# Calibration plot
plot(sales ~ predict(md2), main = "Calibration Plot: sales vs. Prediction", xlab = "Prediction", ylab = "Sales")
abline(0, 1, col = "blue")

# QQ plot of the residuals
qqnorm(resid(md2), main = "Q-Q Plot of the Residuals")
qqline(resid(md2), col = "red")
```

**Statistical explanation**:

- The **residual analysis** serves to check the assumptions of linear regression:
  - Normality: the residuals should be normally distributed.
  - Homoscedasticity: the variance of the residuals should be constant.
  - Independence: the residuals should be independent of one another.
  - Linearity: the relationship between the predictors and the dependent variable should be linear.
- **Residual plots** show the residuals as a function of the predictors or the fitted values.
- A **histogram of the residuals** provides information about the distribution of the residuals.
- A **Q-Q plot** compares the distribution of the residuals with the normal distribution.
- A **calibration plot** shows the observed values as a function of the fitted values. Ideally, the points lie on the diagonal.

**R explanation**:

- The function `resid()` extracts the residuals from a regression model.
- The function `hist()` creates a histogram.
- The function `plot()` with a formula creates a scatterplot.
- The function `abline()` with the parameter `h` adds a horizontal line to the plot.
- The function `qqnorm()` creates a Q-Q plot that compares the distribution of the residuals with the normal distribution.
- The function `qqline()` adds a reference line to the Q-Q plot.

# Extension: Multiple Regression with Three Predictors

We now extend the model with a third predictor: Newspaper.

```{r three-predictors, eval=TRUE}
# Model with TV, Radio and Newspaper
md3 <- lm(sales ~ TV + radio + newspaper)
summary(md3)

# Comparison of R² and RSE
cat("Model comparison:\n")
cat("R² (TV + Radio):", round(summary(md2)$r.squared, 4), "\n")
cat("R² (TV + Radio + Newspaper):", round(summary(md3)$r.squared, 4), "\n")
cat("RSE (TV + Radio):", round(summary(md2)$sigma, 4), "\n")
cat("RSE (TV + Radio + Newspaper):", round(summary(md3)$sigma, 4), "\n\n")

# Adjusted R²
cat("Adjusted R² (TV + Radio):", round(summary(md2)$adj.r.squared, 4), "\n")
cat("Adjusted R² (TV + Radio + Newspaper):", round(summary(md3)$adj.r.squared, 4), "\n")
```

**Statistical explanation**:

- In **multiple regression with three predictors**, the relationship is represented by a hyperplane in four-dimensional space: y = β₀ + β₁x₁ + β₂x₂ + β₃x₃
- The **adjusted R²** takes the number of predictors in the model into account and penalises models with too many predictors:
  - Adj. R² = 1 - [(1 - R²) * (n - 1) / (n - p - 1)]
  - n is the number of observations
  - p is the number of predictors
- The adjusted R² can decrease if an added predictor contributes little to the explanatory power of the model.
- The **residual standard error (RSE)** is a measure of the typical size of the residuals and should be smaller for a better model.
- In this example, adding Newspaper leads to a minimal improvement in R², but the adjusted R² actually decreases slightly, which suggests that Newspaper does not make a substantial contribution to the explanatory power of the model.

**R explanation**:

- The formula `sales ~ TV + radio + newspaper` in `lm()` specifies that `sales` is the dependent variable and `TV`, `radio` and `newspaper` are the independent variables.
- The expression `summary(md2)$adj.r.squared` extracts the adjusted R² from the regression object.
- The expression `summary(md2)$sigma` extracts the RSE from the regression object.

# Non-linear Models: Quadratic and Polynomial Regression

Not all relationships are linear. We now consider non-linear models, in particular quadratic and polynomial regression.

```{r nichtlineare_modelle, eval=TRUE}
# Quadratic regression
md.q2 <- lm(sales ~ TV + I(TV^2) + radio)
summary(md.q2)

# Polynomial regression (degree 3)
md.q3 <- lm(sales ~ poly(TV, 3) + radio)
summary(md.q3)

# Model comparison: Adjusted R²
cat("Model comparison (non-linear):\n")
cat("R² (linear):", round(summary(md2)$adj.r.squared, 4), "\n")
cat("R² (quadratic):", round(summary(md.q2)$adj.r.squared, 4), "\n")
cat("R² (cubic):", round(summary(md.q3)$adj.r.squared, 4), "\n\n")

# Visualisation of the models
# Build a grid of TV values
tv_grid <- seq(min(TV), max(TV), length.out = 100)
radio_mean <- mean(radio)

# Predictions for the various models
pred_linear <- predict(md2, newdata = data.frame(TV = tv_grid, radio = radio_mean))
pred_quad <- predict(md.q2, newdata = data.frame(TV = tv_grid, radio = radio_mean))
pred_poly3 <- predict(md.q3, newdata = data.frame(TV = tv_grid, radio = radio_mean))

# Plot
plot(TV, sales, main = "Comparison of the Models", xlab = "TV", ylab = "Sales")
lines(tv_grid, pred_linear, col = "blue", lwd = 2)
lines(tv_grid, pred_quad, col = "red", lwd = 2, lty = 2)
lines(tv_grid, pred_poly3, col = "green", lwd = 2, lty = 3)
legend("bottomright",
       legend = c("Linear", "Quadratic", "Cubic"),
       col = c("blue", "red", "green"),
       lty = c(1, 2, 3),
       lwd = 2)
```

**Statistical explanation**:

- **Non-linear regression** models non-linear relationships between the predictors and the dependent variable.
- **Quadratic regression** adds a quadratic term: y = β₀ + β₁x + β₂x² + ε
- **Polynomial regression** generalises this to higher powers: y = β₀ + β₁x + β₂x² + β₃x³ + ... + βₚxᵖ + ε
- The function `I(TV^2)` in R adds a quadratic term while preserving the original scale.
- The function `poly(TV, 3)` adds orthogonal polynomials up to degree 3, which is numerically more stable.
- Non-linear models can capture more complex relationships, but they are also more susceptible to overfitting.
- The choice between linear and non-linear models should be based on the nature of the problem, the visualisation of the data, and model comparison criteria such as the adjusted R².

**R explanation**:

- The function `I()` in a formula protects the expression from interpretation by the formula language.
- The function `poly()` produces orthogonal polynomials that are numerically more stable than simple powers.
- The function `seq()` produces a sequence of values.
- The function `lines()` adds lines to an existing plot.
- The parameters `col`, `lwd` and `lty` in `lines()` specify the colour, width and type of the line.

# Model Comparison: R², Adjusted R² and RSE

We now systematically compare the various models on the basis of R², adjusted R² and RSE.

```{r modellvergleich, eval=TRUE}
# Model comparison: R², Adjusted R² and RSE
models <- list(
  "TV" = md1,
  "TV + Radio" = md2,
  "TV + Radio + Newspaper" = md3,
  "TV + TV² + Radio" = md.q2,
  "TV (Poly3) + Radio" = md.q3
)

# Extract R², Adj. R² and RSE for each model
model_stats <- data.frame(
  Model = names(models),
  R2 = sapply(models, function(m) summary(m)$r.squared),
  Adj_R2 = sapply(models, function(m) summary(m)$adj.r.squared),
  RSE = sapply(models, function(m) summary(m)$sigma)
)

# Output the model statistics
# Round only the numeric columns
model_stats_rounded <- model_stats
model_stats_rounded[, 2:4] <- round(model_stats[, 2:4], 4)
print(model_stats_rounded)

# Visualisation of the model comparison
par(mfrow = c(1, 2))
barplot(model_stats$Adj_R2, names.arg = model_stats$Model,
        main = "Adjusted R²", las = 2, cex.names = 0.7,
        col = "lightblue", ylim = c(0, 1))
barplot(model_stats$RSE, names.arg = model_stats$Model,
        main = "Residual Standard Error (RSE)", las = 2, cex.names = 0.7,
        col = "salmon")
par(mfrow = c(1, 1))
```

**Statistical explanation**:

- **Model comparison** is an important step in selecting the best model for the data.
- **R²** gives the proportion of explained variation, but it always increases when predictors are added, even if they are not informative.
- The **adjusted R²** takes the number of predictors into account and penalises models with too many predictors.
- The **residual standard error (RSE)** gives the typical size of the residuals and should be smaller for a better model.
- In model selection one should find a compromise between goodness of fit and model complexity:
  - A model that is too simple may miss important relationships (underfitting).
  - A model that is too complex may model noise and generalise poorly (overfitting).
- In this example, the model "TV + TV² + Radio" appears to offer the best compromise, since it has the highest adjusted R² and a low RSE.

**R explanation**:

- The function `list()` creates a list of objects.
- The function `sapply()` applies a function to each element of a list and returns a vector.
- The function `par(mfrow = c(1, 2))` divides the graphics window into 1 row and 2 columns.
- The function `barplot()` creates a bar chart.
- The parameter `las = 2` in `barplot()` rotates the axis labels.
- The parameter `cex.names = 0.7` in `barplot()` reduces the font size of the names.

# Summary

In this lecture we have introduced multiple linear regression, an extension of simple linear regression to several predictors. The key concepts were:

1. **Multiple regression with several predictors**: multiple regression models the relationship between a dependent variable and several independent variables. The coefficients give the partial effect of each variable when all other variables are held constant.

2. **Residual analysis and goodness of fit**: the analysis of the residuals helps to check the assumptions of linear regression. Goodness of fit is quantified by measures such as R², adjusted R² and RSE.

3. **Prediction and error margin**: with the regression model, predictions can be made for new values of the independent variables. Prediction and confidence intervals quantify the uncertainty of these predictions.

4. **Non-linear and polynomial regression**: not all relationships are linear. Quadratic and polynomial regression can model more complex relationships.

5. **Model comparison**: the choice between different models should be based on criteria such as the adjusted R² and the RSE, with a compromise between goodness of fit and model complexity.

Multiple regression is a powerful tool for data analysis, but it also has limitations. It assumes linear relationships (unless non-linear terms are added explicitly), can be impaired by multicollinearity, and is susceptible to outliers. In such cases, more robust or more flexible methods may be more appropriate.
