---
title: "R Code Examples from WDDA Lecture 01"
subtitle: "WDDA -- Data Management and Data Analysis"
author: "Prof. Dr. Ulrich Matter"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
    latex_engine: xelatex
    md_extensions: "+lists_without_preceding_blankline"
    includes:
      in_header: ../style/notes_preamble.tex
lang: en
---

# Introduction

This document collects all R code examples from the WDDA Lecture 01 slides. It contains additional comments, explanations, and examples to make the individual code snippets easier to understand.


# Importing a Dataset

In this section we import a dataset using the **readxl** package. `read_excel()` automatically detects column types. Use `str()` to inspect the structure of the imported data.

```{r import-dataset, eval=FALSE}
library(readxl)  # Load the readxl package for importing Excel files
markets <- read_excel("WDDA_01.xlsx", sheet = "Exchanges")  # Import the data from the "Exchanges" sheet
```

```{r import-dataset-hidden, echo=FALSE, message=FALSE}
library(readxl)
markets <- read_excel("../../data/WDDA_01.xlsx", sheet = "Exchanges")
```

*Additional tip:* Inspect the structure of the imported data frame:
```{r}
str(markets)  # Show the structure of the data frame 'markets'
```

*Note:* R stores the object `markets` as a data frame.


# R Syntax Styles

R offers several ways of writing code. The two main styles are:

1. **Classic (nested) syntax**: functions are read from the inside out
2. **Pipe syntax**: functions are read from left to right (like a pipeline)

Here is a comparison of the two styles:

```{r syntax-comparison, eval=FALSE}
# Classic syntax (nested)
mean(sqrt(abs(x)))  # Read from inside to outside: first abs(), then sqrt(), then mean()

# Pipe syntax with |>
x |> abs() |> sqrt() |> mean()  # Read left to right: first abs(), then sqrt(), then mean()

# A more complex example:
# Classic
subset(transform(data, z = x + y), z > 0)

# With the pipe
data |>
  transform(z = x + y) |>
  subset(z > 0)
```

The pipe syntax (`|>`) was introduced in R 4.1.0 and is now the recommended style for readable code. The older pipe operator (`%>%`) from the magrittr package works similarly.

*Tip*: In RStudio you can use the keyboard shortcut `Ctrl + Shift + M` to insert the pipe operator. To switch between the native (`|>`) and the magrittr (`%>%`) pipe operator, go to:
Tools -> Global Options -> Code -> "Use native pipe operator"
(Source: [Stack Overflow](https://stackoverflow.com/questions/68667933/how-to-fast-insert-new-pipe-operator-in-rstudio))


# Navigating Data Frames

Here we show different methods for inspecting and subsetting a data frame. Each method is useful for different tasks.

```{r navigation-dataframes, eval=FALSE}
# Show the entire data frame
markets

# Show the column names of the data frame
names(markets)

# Classic subsetting examples:
markets$trades        # Access the column 'trades'
markets[3,]           # Access the third row
markets[,2]           # Access the second column
markets[3,2]          # Access the element in the 3rd row and 2nd column

# Using tidyverse functions for a more readable syntax:
markets |> pull(trades)         # Extract the column 'trades' as a vector
markets |> slice(3)             # Access the third row
markets |> select(2)            # Select the second column by position
markets |> select(trades)       # Select the column 'trades' by name
markets |> slice(2) |> select(trades)   # First take the second row, then select the column 'trades'
```

*Additional example:* Show the first six rows of the data frame:
```{r}
head(markets)  # Show the first six rows of the data frame
```


# Using attach() and detach()

The function `attach()` makes it easier to access the variables of a data frame by adding them to the search path. However, this can lead to naming conflicts if multiple objects share the same name. An alternative is `with()`, which restricts the scope.

```{r attach-detach, eval=FALSE}
# Attach the data frame so that its columns can be accessed directly
attach(markets)

# Now 'trades' can be accessed directly, without prefixing the data frame name
trades

# Remove the data frame from the search path to avoid naming conflicts
detach(markets)

# After detaching, accessing 'trades' without the data frame name will throw an error:
trades
```

*Additional note:* An alternative to `attach()` is `with()`, which only takes effect inside a specific expression:
```{r}
with(markets, summary(trades))  # Summarise the column 'trades' without permanently attaching the data frame
```

# Using length()

The function `length()` behaves differently for data frames and vectors. For a data frame, `length()` returns the number of columns (variables), not the number of rows (observations). To count rows, use `nrow()`.

```{r length-examples, eval=FALSE}
# For a data frame, length() returns the number of columns
markets |> length()

# For a vector, length() returns the number of elements (observations)
trades |> length()

# To get the number of rows (observations) of a data frame, use nrow():
nrow(markets)
```

*Additional tip:* You can use `ncol()` to obtain the number of columns:
```{r}
ncol(markets)
```
