---
title: "WDDA FS 2025: Exercise 1 Guide"
author: "Course Instructor"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

This guide provides detailed explanations for the exercises in WDDA FS 2025 Aufgabenserie 1. For each exercise, we'll walk through the reasoning process and explain how to arrive at the correct solution.

# Exercise 1: Variable Classification

In this exercise, we need to classify variables as nominal, ordinal, or metric (quantitative), and determine if they are discrete or continuous.

## Key Concepts

Before we begin, let's review the key concepts:

- **Nominal variables**: Categories with no inherent order (e.g., colors, gender)
- **Ordinal variables**: Categories with a meaningful order (e.g., rankings, grades)
- **Metric/Quantitative variables**: Numerical measurements that can be:
  - **Discrete**: Countable values (e.g., number of children)
  - **Continuous**: Can take any value within a range (e.g., height, time)

## Solutions with Explanations

Let's go through each variable:

### (a) Number of goals scored weekly by a football team

**Classification**: Metric and discrete

**Explanation**: This is a count variable that can only take whole number values (0, 1, 2, etc.). It's metric because we can perform meaningful arithmetic operations (average goals per game, total goals, etc.), and it's discrete because goals only occur in whole numbers.

### (b) Heights of players in a football team

**Classification**: Metric and continuous

**Explanation**: Height is measured on a continuous scale (e.g., 175.6 cm). It's metric because we can perform arithmetic operations, and it's continuous because height can theoretically take any value within a range, not just whole numbers.

### (c) Most popular radio stations

**Classification**: Nominal and discrete

**Explanation**: Radio stations are categories with no inherent order. One station isn't "more" than another in any quantitative sense. It's discrete because we're categorizing distinct entities.

### (d) Number of children in a Japanese family

**Classification**: Metric and discrete

**Explanation**: This is a count variable that can only take whole number values. It's metric because we can calculate meaningful statistics like averages, and it's discrete because people only come in whole numbers.

### (e) Military rank of a soldier

**Classification**: Ordinal and discrete

**Explanation**: Military ranks have a clear hierarchical order (e.g., private, corporal, sergeant). It's ordinal because the ranks have a meaningful sequence, and it's discrete because there are distinct categories.

### (f) Number of loaves of bread a family buys per week

**Classification**: Metric and discrete

**Explanation**: This is a count variable that can only take whole number values. It's metric because we can perform arithmetic operations, and it's discrete because bread loaves are purchased as whole units.

### (g) Types of pets owned by 8th grade students

**Classification**: Nominal and discrete

**Explanation**: Pet types (dog, cat, fish, etc.) are categories with no inherent order. It's nominal because these categories don't have a natural ranking, and it's discrete because we're dealing with distinct categories.

### (h) Number of leaves on trees

**Classification**: Metric and discrete

**Explanation**: This is a count variable that can only take whole number values. It's metric because we can perform arithmetic operations, and it's discrete because leaves are counted as whole units.

### (i) Number of sunshine hours in a day

**Classification**: Metric and continuous

**Explanation**: Sunshine hours can be measured with precision beyond whole numbers (e.g., 5.7 hours). It's metric because we can perform arithmetic operations, and it's continuous because time can be measured on a continuous scale.

### (j) Number of people who die from cancer each year in the USA

**Classification**: Metric and discrete

**Explanation**: This is a count variable that can only take whole number values. It's metric because we can perform arithmetic operations, and it's discrete because we're counting individual people.

### (k) Countries of origin of immigrants

**Classification**: Nominal and discrete

**Explanation**: Countries are categories with no inherent numerical order. It's nominal because countries don't have a natural ranking, and it's discrete because we're dealing with distinct categories.

### (l) Colors of cars

**Classification**: Nominal and discrete

**Explanation**: Colors are categories with no inherent order. It's nominal because colors don't have a natural ranking, and it's discrete because we're dealing with distinct categories.

### (m) Gender of school principals

**Classification**: Nominal and discrete

**Explanation**: Gender categories have no inherent order. It's nominal because gender doesn't have a natural ranking, and it's discrete because we're dealing with distinct categories.

### (n) Time spent on homework

**Classification**: Metric and continuous

**Explanation**: Time can be measured with precision beyond whole numbers (e.g., 2.5 hours). It's metric because we can perform arithmetic operations, and it's continuous because time can be measured on a continuous scale.

### (o) Categories of poker hands

**Classification**: Ordinal and discrete

**Explanation**: Poker hands have a clear ranking (high card, pair, two pair, etc.). It's ordinal because there's a meaningful order to the categories, and it's discrete because there are distinct categories.

### (p) Amount spent on groceries in dollars per week

**Classification**: Metric and continuous

**Explanation**: Money can be measured with precision beyond whole numbers (e.g., $45.67). It's metric because we can perform arithmetic operations, and it's continuous because money can be measured on a continuous scale.

### (q) Grades achieved in a class test

**Classification**: Ordinal and discrete

**Explanation**: Grades (A, B, C, etc. or numerical grades) have a clear ranking. It's ordinal because there's a meaningful order to the categories, and it's discrete because there are distinct grade levels.

### (r) Number of items sold in the school bistro

**Classification**: Metric and discrete

**Explanation**: This is a count variable that can only take whole number values. It's metric because we can perform arithmetic operations, and it's discrete because items are sold as whole units.

### (s) Number of matches in a box

**Classification**: Metric and discrete

**Explanation**: This is a count variable that can only take whole number values. It's metric because we can perform arithmetic operations, and it's discrete because matches are counted as whole units.

### (t) Final positions of athletes in a race

**Classification**: Ordinal and discrete

**Explanation**: Race positions (1st, 2nd, 3rd, etc.) have a clear ranking. It's ordinal because there's a meaningful order to the positions, and it's discrete because there are distinct positions.

### (u) Final times of athletes in a race

**Classification**: Metric and continuous

**Explanation**: Race times can be measured with precision beyond whole numbers (e.g., 10.45 seconds). It's metric because we can perform arithmetic operations, and it's continuous because time can be measured on a continuous scale.

### (v) Reasons why people use taxis

**Classification**: Nominal and discrete

**Explanation**: Reasons (convenience, no car, etc.) are categories with no inherent order. It's nominal because these reasons don't have a natural ranking, and it's discrete because we're dealing with distinct categories.

### (w) Sports played by students at high schools

**Classification**: Nominal and discrete

**Explanation**: Sports (basketball, soccer, etc.) are categories with no inherent order. It's nominal because sports don't have a natural ranking, and it's discrete because we're dealing with distinct categories.

### (x) Consumer preference between three different products

**Classification**: Nominal and discrete

**Explanation**: Product preferences are categories with no inherent order. It's nominal because preferences don't necessarily have a natural ranking, and it's discrete because we're dealing with distinct categories.

### (y) Pulse rates of a group of athletes at rest

**Classification**: Metric and continuous

**Explanation**: Pulse rates can be measured with precision beyond whole numbers (e.g., 72.5 beats per minute). It's metric because we can perform arithmetic operations, and it's continuous because heart rates can vary continuously.

# Exercise 2: Determining Variable Values

This exercise asks us to discuss how we might determine values for various variables.

## (a) Where a person stands on the political spectrum

To determine where a person stands on the political spectrum, we could:

1. **Use a survey instrument**: Create a questionnaire with statements about political issues, asking respondents to indicate their level of agreement on a scale (e.g., 1-7 Likert scale).

2. **Analyze voting history**: Look at past voting patterns to place individuals on a spectrum.

3. **Self-identification**: Simply ask people to place themselves on a political spectrum (e.g., "On a scale from 1 (very liberal) to 10 (very conservative), where would you place yourself?").

4. **Issue-based mapping**: Ask about positions on specific issues and map responses to a political spectrum.

The resulting variable would likely be ordinal (if using categories like "very liberal," "somewhat liberal," etc.) or potentially metric if using a numerical scale.

## (b) A patient's blood pressure

Blood pressure is typically determined by:

1. **Direct measurement**: Using a sphygmomanometer (blood pressure cuff) and stethoscope or an automated blood pressure monitor.

2. **Multiple readings**: Taking several measurements at different times to account for variation.

3. **Standardized conditions**: Ensuring the patient is seated, relaxed, arm at heart level, etc.

Blood pressure is recorded as two numbers: systolic (pressure when heart beats) and diastolic (pressure when heart rests). For example, 120/80 mmHg.

This is a metric, continuous variable as it can take any value within a range and can be measured with high precision.

## (c) A country's poverty line

A country's poverty line can be determined through:

1. **Absolute approach**: Setting a minimum income needed for basic necessities (food, shelter, clothing).

2. **Relative approach**: Defining poverty as a percentage of median income (e.g., 60% of median income).

3. **Multidimensional approach**: Considering factors beyond income, such as access to education, healthcare, and basic services.

4. **Consumption-based measures**: Looking at what households consume rather than their income.

The resulting variable is metric and continuous, typically expressed as a monetary value per person or household per time period.

## (d) An athlete's endurance performance

An athlete's endurance performance can be measured through:

1. **Time-to-exhaustion tests**: How long an athlete can maintain a specific intensity of exercise.

2. **Distance tests**: How far an athlete can go in a set time (e.g., 12-minute run test).

3. **VO2 max testing**: Measuring maximum oxygen consumption during incremental exercise.

4. **Heart rate recovery**: How quickly heart rate returns to normal after exertion.

5. **Lactate threshold testing**: Determining the exercise intensity at which lactate begins to accumulate in the blood.

This is a metric variable that could be continuous (if measuring time or physiological parameters) or discrete (if using standardized performance categories).

## (e) The number of times a person lies in a day

This is challenging to measure accurately, but approaches might include:

1. **Self-reporting**: Asking people to keep a diary of when they lie, though this is subject to honesty and awareness issues.

2. **Observation**: Having trained observers monitor interactions, though this is resource-intensive and may miss subtle lies.

3. **Experimental settings**: Creating situations where lying might occur and monitoring behavior.

4. **Physiological measures**: Using polygraph-type measurements in controlled settings.

This would be a metric, discrete variable as lies would be counted as whole numbers.

# Exercise 3: Graph Analysis

In this exercise, we analyze a graph comparing MSNBC and Fox News.

## (a) Identify the two variables

The two variables shown in the graph are:

1. **Factual Reporting**: Represented by the top bar, ranging from "Very Low" to "Very High"
2. **Political Bias**: Represented by the bottom arrow, ranging from "Extreme Left" to "Extreme Right"

## (b) Explain how values for these variables could be determined

### Factual Reporting

Values for factual reporting could be determined by:

1. **Content analysis**: Examining a sample of news stories and fact-checking claims against primary sources or established facts.

2. **Error rate tracking**: Counting corrections, retractions, and identified false claims over time.

3. **Expert panel evaluation**: Having media experts or fact-checkers rate content based on established criteria.

4. **Source transparency**: Evaluating whether claims are properly attributed to verifiable sources.

5. **Methodology disclosure**: Assessing whether the news outlet explains how information was gathered.

### Political Bias

Values for political bias could be determined by:

1. **Content analysis**: Examining language, framing, and topic selection for partisan leanings.

2. **Source selection**: Analyzing which experts, politicians, or organizations are quoted or featured.

3. **Issue coverage**: Measuring time/space devoted to different political issues.

4. **Audience surveys**: Asking viewers/readers about their own political leanings to identify patterns.

5. **Comparative analysis**: Comparing coverage of the same events across different outlets.

## (c) What type are the variables?

Both variables are presented as **ordinal** and **discrete** variables in the graph.

**Explanation**:
- They are ordinal because the categories have a meaningful order (from "Very Low" to "Very High" for factual reporting, and from "Extreme Left" to "Extreme Right" for political bias).
- They are presented as discrete because they use distinct categories rather than a continuous scale.

However, it's worth noting that the underlying data collection might have used more continuous measurements that were then converted to these ordinal categories for presentation purposes.

# Exercises 4-7: Working with the BFH Dataset

These exercises involve working with variables from the BFH dataset and performing operations on them.

## Exercise 4: sum(cash) vs sum(gender)

```{r, eval=FALSE}
# This is conceptual - we don't have the actual BFH dataset loaded
sum(cash)   # Would return the total amount of cash carried by all individuals
sum(gender) # Would return an error because gender is a nominal variable
```

**Explanation**:
- `sum(cash)` works because `cash` is a metric variable (amount of CHF carried). Adding these values gives us the total amount of cash carried by all individuals in the dataset.
- `sum(gender)` produces an error because `gender` is a nominal variable (categories with no inherent numerical value). R cannot perform arithmetic operations on non-numeric data.

## Exercise 5: height/100

```{r, eval=FALSE}
# This is conceptual - we don't have the actual BFH dataset loaded
height/100  # Converts height from centimeters to meters
```

**Explanation**:
This operation converts the height measurements from centimeters to meters. Since the original `height` variable is in centimeters, dividing by 100 gives us the equivalent height in meters.

For example, if someone is 175 cm tall, height/100 would give 1.75 m.

## Exercise 6: reaction1-reaction2

```{r, eval=FALSE}
# This is conceptual - we don't have the actual BFH dataset loaded
reaction1-reaction2  # Calculates the difference in reaction times between first and second attempts
```

**Explanation**:
This operation calculates the difference between the first and second reaction time measurements. 

- If the result is positive (reaction1 > reaction2), it indicates an improvement in reaction time from the first to the second attempt.
- If the result is negative (reaction1 < reaction2), it indicates a slower reaction time in the second attempt.
- If the result is zero, it indicates no change in reaction time.

This new variable represents the change or improvement in reaction time between attempts.

## Exercise 7: siblings*present

```{r, eval=FALSE}
# This is conceptual - we don't have the actual BFH dataset loaded
siblings*present  # Calculates total gift expenditure per person
sum(siblings*present)  # Would calculate total gift expenditure across all people
```

**Explanation**:
The operation `siblings*present` multiplies the number of siblings a person has by the amount spent on gifts per sibling. This gives us the total amount each person spends on gifts for all their siblings.

The sum of this variable across all individuals (`sum(siblings*present)`) would represent the total amount spent on sibling gifts by everyone in the dataset.

# Exercises 8-16: Working with Summation Notation

These exercises focus on understanding and manipulating summation notation.

## Exercise 8: Writing out sums

### (a) $\sum_{r=1}^{4} r^3$

This means: $1^3 + 2^3 + 3^3 + 4^3$

Calculating: $1 + 8 + 27 + 64 = 100$

### (b) $\sum_{r=1}^{6} (2r + 1)$

This means: $(2 \times 1+1) + (2 \times 2+1) + (2 \times 3+1) + (2 \times 4+1) + (2 \times 5+1) + (2 \times 6+1)$

Calculating: $3 + 5 + 7 + 9 + 11 + 13 = 48$

### (c) $\sum_{n=1}^{5} 3^{2n}$

This means: $3^{2 \times 1} + 3^{2 \times 2} + 3^{2 \times 3} + 3^{2 \times 4} + 3^{2 \times 5}$

Calculating: $3^2 + 3^4 + 3^6 + 3^8 + 3^{10}$

Calculating: $9 + 81 + 729 + 6,561 + 59,049 = 66,429$

### (d) $\sum_{n=1}^{7} 4$

This means: $4 + 4 + 4 + 4 + 4 + 4 + 4$

Calculating: $7 \times 4 = 28$

## Exercise 9: Using sigma notation for sums

### (a) $1^2 + 2^2 + 3^2 + ... + 7^2$

This can be written as: $\sum_{n=1}^{7} n^2$

### (b) $2^1 + 2^2 + 2^3 + ... + 2^{10}$

This can be written as: $\sum_{n=1}^{10} 2^n$

### (c) $\frac{1}{1} + \frac{1}{2} + \frac{1}{3} + ... + \frac{1}{77}$

This can be written as: $\sum_{n=1}^{77} \frac{1}{n}$

### (d) $\frac{1 \times 3}{2 \times 4} + \frac{2 \times 4}{3 \times 5} + ... + \frac{19 \times 21}{20 \times 22}$

This can be written as: $\sum_{n=1}^{19} \frac{n \times (n+2)}{(n+1) \times (n+3)}$

## Exercise 10: Evaluating sums

### (a) $\sum_{r=1}^{6} r$

This means: $1 + 2 + 3 + 4 + 5 + 6 = 21$

### (b) $\sum_{r=1}^{6} 3r$

This means: $3 \times 1 + 3 \times 2 + 3 \times 3 + 3 \times 4 + 3 \times 5 + 3 \times 6$

Calculating: $3 + 6 + 9 + 12 + 15 + 18 = 63$

### (c) $3 \times \sum_{r=1}^{6} r$

This means: $3 \times (1 + 2 + 3 + 4 + 5 + 6) = 3 \times 21 = 63$

Note that (b) and (c) give the same result, demonstrating the property that a constant can be factored out of a sum.

## Exercise 11: Writing out general sums

### (a) $\sum_{r=1}^{n} x_r$

This means: $x_1 + x_2 + x_3 + ... + x_n$

### (b) $\sum_{r=1}^{n} x_r^2$

This means: $x_1^2 + x_2^2 + x_3^2 + ... + x_n^2$

### (c) $\sum_{r=1}^{n} x_r^r$

This means: $x_1^1 + x_2^2 + x_3^3 + ... + x_n^n$

## Exercise 12: True or false statements

### (a) $\sum_{r=1}^{n} x_r + \sum_{r=1}^{n} y_r = \sum_{r=1}^{n} (x_r + y_r)$

**True**

Explanation: This property states that the sum of two separate summations equals the summation of the sum of corresponding terms. This is true because addition is commutative and associative.

For example, with n=3:
$(x_1 + x_2 + x_3) + (y_1 + y_2 + y_3) = (x_1 + y_1) + (x_2 + y_2) + (x_3 + y_3)$

### (b) $\sum_{r=1}^{n} x_r \cdot \sum_{r=1}^{n} y_r = \sum_{r=1}^{n} (x_r \cdot y_r)$

**False**

Explanation: The left side multiplies every term in the first sum by every term in the second sum, while the right side only multiplies corresponding terms.

For example, with n=2:
$(x_1 + x_2) \cdot (y_1 + y_2) = x_1y_1 + x_1y_2 + x_2y_1 + x_2y_2$
But $\sum_{r=1}^{2} (x_r \cdot y_r) = x_1y_1 + x_2y_2$

These are clearly not equal.

### (c) $\sum_{r=1}^{n} (c \cdot x_r) = c \cdot \sum_{r=1}^{n} x_r$

**True**

Explanation: This property states that a constant can be factored out of a summation. This is true because multiplication distributes over addition.

For example, with n=3:
$c \cdot x_1 + c \cdot x_2 + c \cdot x_3 = c \cdot (x_1 + x_2 + x_3)$

## Exercise 13: Writing out $\sum_{m=1}^{4} (-1)^m x$

This means: $(-1)^1 \cdot x + (-1)^2 \cdot x + (-1)^3 \cdot x + (-1)^4 \cdot x$

Calculating: $-x + x - x + x = 0$

## Exercise 14: Finding $\sum_{r=1}^{7} (x_r - 2)^2$ given other sums

Given:
- $\sum_{r=1}^{7} x_r^2 = 100$
- $\sum_{r=1}^{7} x_r = 10$

We need to find: $\sum_{r=1}^{7} (x_r - 2)^2$

Let's expand:
$\sum_{r=1}^{7} (x_r - 2)^2 = \sum_{r=1}^{7} (x_r^2 - 4x_r + 4)$

$= \sum_{r=1}^{7} x_r^2 - 4\sum_{r=1}^{7} x_r + \sum_{r=1}^{7} 4$

$= \sum_{r=1}^{7} x_r^2 - 4\sum_{r=1}^{7} x_r + 4 \cdot 7$

$= 100 - 4 \cdot 10 + 28$

$= 100 - 40 + 28$

$= 88$

## Exercise 15: Writing out various statistical notations

### (a) $\frac{1}{N} \cdot \sum_{i=1}^{N} x_i$

This is the formula for the arithmetic mean (average) of a set of values:
$\frac{x_1 + x_2 + x_3 + ... + x_N}{N}$

### (b) $\sum_{r=1}^{n} (x_r - \bar{x})^2$

This is the sum of squared deviations from the mean, used in calculating variance:
$(x_1 - \bar{x})^2 + (x_2 - \bar{x})^2 + (x_3 - \bar{x})^2 + ... + (x_n - \bar{x})^2$

### (c) $\sum_{i=1}^{n} w_i x_i$

This is a weighted sum, where each value $x_i$ is multiplied by a weight $w_i$:
$w_1 x_1 + w_2 x_2 + w_3 x_3 + ... + w_n x_n$

### (d) $\sum_{i=1}^{n} P(E_i)$

This is the sum of probabilities for events $E_1$ through $E_n$:
$P(E_1) + P(E_2) + P(E_3) + ... + P(E_n)$

### (e) $\sum_{i=1}^{3} \sum_{j=1}^{2} P_{ij}$

This is a double summation, summing values in a 3×2 matrix:
$P_{11} + P_{12} + P_{21} + P_{22} + P_{31} + P_{32}$

### (f) $\sum_{i=1}^{n} (x_i - \bar{x})(y_i - \bar{y})$

This is the sum of products of deviations, used in calculating covariance:
$(x_1 - \bar{x})(y_1 - \bar{y}) + (x_2 - \bar{x})(y_2 - \bar{y}) + ... + (x_n - \bar{x})(y_n - \bar{y})$

## Exercise 16: Evaluating table-based summations

Given the table:
```
  1  2  3  4
  5  6  7  8
  9 10 11 12
 13 14 15 16
```

Where a(r,c) is the value in row r, column c.

### (a) $\sum_{n=1}^{4} a(n, 2)$

This means: a(1,2) + a(2,2) + a(3,2) + a(4,2)

Looking at column 2: 2 + 6 + 10 + 14 = 32

### (b) $\sum_{n=1}^{4} a(1, n)$

This means: a(1,1) + a(1,2) + a(1,3) + a(1,4)

Looking at row 1: 1 + 2 + 3 + 4 = 10

### (c) $\sum_{n=1}^{4} a(n, n)$

This means: a(1,1) + a(2,2) + a(3,3) + a(4,4)

Looking at the diagonal: 1 + 6 + 11 + 16 = 34

### (d) $\sum_{n=1}^{4} (a(n, 1) \times a(n, 3))$

This means: a(1,1)$\times$a(1,3) + a(2,1)$\times$a(2,3) + a(3,1)$\times$a(3,3) + a(4,1)$\times$a(4,3)

Calculating: 1$\times$3 + 5$\times$7 + 9$\times$11 + 13$\times$15 = 3 + 35 + 99 + 195 = 332

### (e) $\sum_{n=1}^{4} a(n, 1) \times \sum_{n=1}^{4} a(n, 3)$

This means: (a(1,1) + a(2,1) + a(3,1) + a(4,1)) $\times$ (a(1,3) + a(2,3) + a(3,3) + a(4,3))

Calculating: (1 + 5 + 9 + 13) $\times$ (3 + 7 + 11 + 15) = 28 $\times$ 36 = 1008

### (f) $\sum_{n=1}^{4} a(n, 3) + \sum_{m=1}^{4} a(4, m)$

This means: (a(1,3) + a(2,3) + a(3,3) + a(4,3)) + (a(4,1) + a(4,2) + a(4,3) + a(4,4))

Calculating: (3 + 7 + 11 + 15) + (13 + 14 + 15 + 16) = 36 + 58 = 94

### (g) $\sum_{n=1}^{4} \sum_{m=1}^{4} a(n, m)$

This means the sum of all values in the table:
1 + 2 + 3 + 4 + 5 + 6 + 7 + 8 + 9 + 10 + 11 + 12 + 13 + 14 + 15 + 16 = 136

### (h) $\sum_{n=2}^{4} \sum_{m=1}^{n-1} a(n, m)$

This is a bit more complex:

For n=2: m goes from 1 to 1, so we have a(2,1) = 5
For n=3: m goes from 1 to 2, so we have a(3,1) + a(3,2) = 9 + 10 = 19
For n=4: m goes from 1 to 3, so we have a(4,1) + a(4,2) + a(4,3) = 13 + 14 + 15 = 42

Total: 5 + 19 + 42 = 66

# Conclusion

This guide has provided detailed explanations for all the exercises in WDDA FS 2025 Aufgabenserie 1. By working through these exercises, you should have developed a better understanding of:

1. How to classify variables by type (nominal, ordinal, metric) and nature (discrete, continuous)
2. Methods for determining values for different types of variables
3. How to analyze graphical representations of data
4. Working with dataset variables and creating derived variables
5. Understanding and manipulating summation notation
6. Applying summation properties to solve problems

These foundational concepts are essential for more advanced data analysis techniques that will be covered in future exercises.
